Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-06-15 manu_draft.Rmd
log chl models added from sfbaytrends, minor edits to methods

Nitrogen and phospohorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.  However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.  The resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.  Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999 (Figure \@ref(fig:obsdat)).  These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.  Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, how do current conditions compare to desirable water quality conditions, and what potential management actions are needed to achieve desired conditions if they currently do not persist.  Although visual changes in observed data were apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed.

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20] (Figure \@ref(fig:sitemap)).  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother [@Wood03].  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.

An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location. Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is a useful form for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.  A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of component being modelled.  This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.  In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.  This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences over time.

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM perforamcne between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02 to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

@@

where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.  This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.

@@

>>>>

Nitrogen and phospohorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.  However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.  The resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.  Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999 (Figure \@ref(fig:obsdat)).  These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.  Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, describing how current conditions compare to desirable water quality conditions, and identifying what potential management actions are needed to achieve desired conditions if they currently do not persist.  Although visual changes in observed data were apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed.

The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].  This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.  Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients. The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)). Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters. Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.  While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).

@@

The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.  A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.  Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.  There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother [@Wood03].  The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).  In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty. This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function. However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.  The effect of this choose on model fit is described below.

An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location. Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is useful for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.  A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of a component being modelled.  This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.  In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.  This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences.

@@

The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.  A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.  The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable. However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.  The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).  The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.  The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.  Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:

@@

where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value. Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.  The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.  Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.

A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.  The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).  However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.  Initial testing by Murphy et al. (2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.  Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.   We follow the same approach herein for gam1 and gam2.  Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further. For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.  This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.

@@

* start with description of model performance, which models had the best performance (gam0 v gam6, boxcox v log)
* then describe figure 4
